Predicting FOMC Actions using NLP; Assesing the impact of matrix sparsity and regularization

### 2019-05-14 presentation/paperv3.tex
edits to paper v3
update paper

\subsection{Matrix Sizes and Sparsity from CountVector}

@@

As we mentioned in our introduction, the classic problem with NLP bag-of-words models is that most of the column entries (n-grams) are zero, resulting in extremly sparse matricies.  The table below gives a measure of $\mbox{sparsity} = \frac{non-zero elements}{total elements}$ and the resulting number of elements in each n-gram generated matrix (units are in billions).  We empirically observed that the ``stacked'' model reduces the sparsity by a factor of four and the matrix size by a factor of five. The L1 matrix norm for all of the matricies ranged between 1200 to 1600.\\

@@

\subsubsection{A comparision of Multinomial Naive Bayes to Logistic Regression with extrem sparse Matricies}
We noticed that under extreme levels of sparseness (sparseness measure of less than $0.003$) from the Matricies 
generated by CountVectorizer , Logistic Regression and Logistic Lasso with larger regularization parameters would break down, 
while Multinomial Naive Bayes would produce usable results. Logistic Lasso with $\alpha=0.002$ seemed to hold up
well even under extreme sparsness.

\section{Conclusion}
Reducing the sparsity of the matricies generated from CountVectorizer by stacking the documents greatly improved the results.
We were able to attain maximum test accuracy of 95.9$\%$ using a 10:15 n-gram with a recall of 93.4$\%$ and an F1 score of 92.5$\%$.
Because of the number of features generated can be in the millions, Logistic Lasso with small penatly proved to be the most effective 
method for reducing the number of coefficients.
The fraction of non-zero coefficients remaining after applying logistic lasso
ranged between 3e-05 to 1.6e-03.  (Compared to .97 with no regularization using logistic regression).
Logistic regression ($\alpha > 0.05$) and Logistic Lasso both broke down at extreme levels of sparsity 
(2.5e-03).
In general the Regularization parameter at the extremes produced inferior results.
MultiNomial Naive Bayes with Laplace smoothing at zero proved to be the most stable method for producing useful results at the
most extreme measures of matrix sparcity.  In general no smoothing or large smoothing proved to give the best results with
MultiNomial Naive Bayes. Overall Logistic Lasso with $\alpha = 0.002$ produced the most stable results.

@@

\subsubsection{Data Sources}
The time frame of our study covered 2000 to 2018.  
The primary source of our textual data were FOMC statements, FOMC meeting minutes, and Board of Governors speeches from \url{https://www.federalreserve.gov}.  
For FOMC statements between 2000 and 2008 we utilized Stanford's FOMC corpus.  
The Stanford FOMC corpus only has text prior to 2008 and it does not have Board of Governor speeches.
The data for the Federal Funds target rate and target range (post 2009) is from FRED St. Louis.  
FRED offers a wealth of economic data and information to promote economic education and enhance economic research. 
FRED is updated regularly and allows 24/7 access to regional and national financial and economic data.

>>>>

\subsection{Matrix Sizes and Sparsity from CountVectorizer}

@@

As we mentioned in our introduction, the classic problem with NLP bag-of-words models is that most of the column entries (n-grams) are zero, resulting in extremly sparse matricies.  
Sparsity is defined as
\begin{equation*}
\mbox{sparsity} = \frac{\mbox{non-zero elements}}{\mbox{total elements}}
\end{equation*}

The table below gives a measure of sparsity  and the resulting number of elements in each n-gram generated matrix (units are in billions).  We empirically observed that the ``stacked'' model reduces the sparsity by a factor of four and the matrix size by a factor of five. The L1 matrix norm for all of the matricies ranged between 1200 to 1600.\\

@@

\subsubsection{A comparision of Multinomial Naive Bayes to Logistic Regression with extrem sparse Matricies}
Under extreme levels of sparseness (sparseness measure of less than $0.0021$) from  Matricies 
generated by CountVectorizer , Logistic Regression and Logistic Lasso break down while
while Multinomial Naive Bayes can still produce usable results. Logistic Lasso with relaitvely low penatly 
$\alpha=0.002$ functions as well or better than Naive Bayes when sparseness exceeded $0.0025$.
Logistic Lasso with a larger penalty $> 2.0$ does not performe will with extremley sparse matricies. \\

\noindent \begin{tabular}{ |p{2cm}||p{2cm}|p{2cm}|p{2cm}|p{2cm}|  }
 \hline

 \multicolumn{5}{|c|}{Deterioration of Logistic Lasso with sparse matricies relative to Naive Bayes} \\
 \hline
 \multicolumn{1}{|c|}{} &
 \multicolumn{1}{|c|}{} &
 \multicolumn{3}{|c|}{F1 Score} \\

 \hline
 \multicolumn{1}{|c|}{} &
 \multicolumn{1}{|c|}{} &
 \multicolumn{1}{|c|}{Naive Bayes} &
 \multicolumn{2}{|c|}{Logistic Lasso} \\

 \hline
 \multicolumn{1}{|c|}{N-gram} &
 \multicolumn{1}{|c|}{sparsity} &
 \multicolumn{1}{|c|}{smooth$=0.0$} &
 \multicolumn{1}{|c|}{$\alpha=0.002$} &
 \multicolumn{1}{|c|}{$\alpha=1.000$} \\
 \hline
 60& 0.001979 & 0.647 & 0.463 & 0.256\\
 50& 0.002018 & 0.619 & 0.565 & 0.342\\
 30& 0.002038 & 0.724 & 0.620 & 0.460\\
 25& 0.002063 & 0.720 & 0.664 & 0.490\\
 20& 0.002067 & 0.725 & 0.678 & 0.520\\
 12& 0.002069 & 0.682 & 0.743 & 0.656\\
 \hline
\end{tabular}

\section{Conclusion}
Reducing the sparsity of the matricies generated from CountVectorizer by stacking the documents greatly improved the results.
We were able to attain maximum test accuracy of 95.9$\%$ using a 10:15 n-gram with a recall of 93.4$\%$ and an F1 score of 92.5$\%$.
Because of the number of features generated can be in the millions, Logistic Lasso with small penatly proved to be the most effective 
method for reducing the number of coefficients.
The fraction of non-zero coefficients remaining after applying logistic lasso
ranged between 3e-05 to 1.6e-03.  (Compared to .97 with no regularization using logistic regression).
Logistic regression  and Logistic Lasso both broke down at the most extreme levels of sparsity $<0.0020$ 
In general Regularization parameters at the extremes ($\alpha=0.0$ or $\alpha > 2.0$) produced inferior results.
MultiNomial Naive Bayes with Laplace smoothing$=0$ was the most stable method for producing useful results at the
most extreme measures of matrix sparcity.  In general no smoothing or large smoothing proved to give the best results with
MultiNomial Naive Bayes. Overall Logistic Lasso with $\alpha = 0.002$ produced the most stable results.

@@

\subsubsection{Data Sources}
The time frame of our study covered 2000 to 2018.  
The primary source of our textual data were FOMC statements, FOMC meeting minutes, and Board of Governors speeches from \url{https://www.federalreserve.gov}.  
The data for the Federal Funds target rate and target range (post 2009) is from FRED St. Louis.  
FRED offers a wealth of economic data and information to promote economic education and enhance economic research. 
FRED is updated regularly and allows 24/7 access to regional and national financial and economic data.
